---
title: "NF1 patients"
subtitle: "Tumor cells dataset"
author: "Audrey"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    code_folding: show
    code_download: true
    toc: true
    toc_float: true
    number_sections: false
---

body {
text-align: justify}
</style>

<!-- Automatically computes and prints in the output the running time for any code chunk -->
```{r, echo=FALSE}
# https://github.com/rstudio/rmarkdown/issues/1453
hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    res = hooks[[type]](x, options)
    
    if (isFALSE(options[[paste0("fold_", type)]])) return(res)
    
    paste0(
      "<details><summary>", "show", "</summary>\n\n",
      res,
      "\n\n</details>"
    )
  }
}
knitr::knit_hooks$set(
  output = hook_foldable("output"),
  plot = hook_foldable("plot"),
  time_it = local({
    now = NULL
    function(before, options) {
      if (options$time_it) {
        if (before) {
          now <<- Sys.time()
        } else {
          res = difftime(Sys.time(), now, units = "secs")
          paste("(Time to run :", round(res, digits = 2), "s)")
        }
      }
    }
  })
)
```

<!-- Set default parameters for all chunks -->
```{r, setup, include = FALSE}
set.seed(1337L)
knitr::opts_chunk$set(echo = TRUE, # display code
                      # display chunk output
                      message = FALSE,
                      warning = FALSE,
                      fold_output = FALSE, # usefull for sessionInfo()
                      fold_plot = FALSE,
                      
                      # figure settings
                      fig.align = 'center',
                      fig.width = 20,
                      fig.height = 15,
                      
                      # something about seed, chunk and Rmarkdown compilation
                      # https://stackoverflow.com/questions/39417003/long-vectors-not-supported-yet-error-in-rmd-but-not-in-r-script
                      # cache = TRUE,
                      cache.lazy = FALSE, 
                      
                      # add runtime after chunk
                      time_it = FALSE)
```

This script is used to generate a tumor cells dataset. Tumor cells and fibroblasts are difficult to distinguish. We extract cells from the combined dataset. Merging cells together helps to make difference between both cell types.

```{r library}
library(patchwork)
library(ggplot2)
library(dplyr)

.libPaths()
```

# Preparation

In this section, we set the global settings of the analysis. We will store data there :

```{r out_dir}
save_name = "patients_tumor_cells"
out_dir = "."
n_threads = 3 # Rtsne::Rtsne option
```

We load the sample information :

```{r custom_palette_sample, fig.width = 8, fig.height = 8}
sample_info = readRDS(paste0(out_dir, "/../1_metadata/patients_sample_info.rds"))
project_names_oi = sample_info$project_name

graphics::pie(rep(1, nrow(sample_info)),
              col = sample_info$color,
              labels = sample_info$project_name)
```

Here are custom colors for each cell type :

```{r color_markers, fig.width = 10, fig.height = 1, class.source = "fold-hide"}
color_markers = readRDS(paste0(out_dir, "/../1_metadata/patients_color_markers.rds"))

data.frame(cell_type = names(color_markers),
           color = unlist(color_markers)) %>%
  ggplot2::ggplot(., aes(x = cell_type, y = 0, fill = cell_type)) +
  ggplot2::geom_point(pch = 21, size = 5) +
  ggplot2::scale_fill_manual(values = unlist(color_markers), breaks = names(color_markers)) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "none",
                 axis.line = element_blank(),
                 axis.title = element_blank(),
                 axis.ticks = element_blank(),
                 axis.text.y = element_blank())
```

# Make `r save_name` dataset

## Load combined dataset

We load the dataset containing all cells from all samples :

```{r load_sobj}
sobj = readRDS(paste0(out_dir, "/../3_combined/all_patients_sobj.rds"))
sobj
```

We smooth cell type annotation at a cluster level :

```{r smooth_annotation}
cluster_type = table(sobj$cell_type, sobj$seurat_clusters) %>%
  prop.table(., margin = 2) %>%
  apply(., 2, which.max)
cluster_type = setNames(nm = names(cluster_type),
                        levels(sobj$cell_type)[cluster_type])

sobj$cluster_type = cluster_type[sobj$seurat_clusters]

sobj$cluster_type = factor(sobj$cluster_type, levels = levels(sobj$cell_type))

table(sobj$cluster_type, sobj$cell_type)
```

We visualize cell types, clustering and tumor type on the tSNE :

```{r see_cell_type_clusters, fig.width = 10, fig.height = 10}
name2D = "harmony_80_tsne"

plot_list = list()

# Tumor type
plot_list[[1]] = Seurat::DimPlot(sobj, group.by = "tumor_type",
                                 reduction = name2D,
                                 cols = c("blue", "green", "brown")) +
  ggplot2::labs(title = "Tumor type") +
  Seurat::NoAxes() +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5),
                 legend.position = "bottom")

# Cell type
plot_list[[2]] = Seurat::DimPlot(sobj, group.by = "cell_type",
                                 reduction = name2D,
                                 cols = color_markers) +
  ggplot2::labs(title = "Cell type") +
  Seurat::NoAxes() +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5),
                 legend.position = "bottom")

# Cluster
plot_list[[3]] = Seurat::DimPlot(sobj, group.by = "seurat_clusters",
                                 reduction = name2D,
                                 label = TRUE) +
  ggplot2::labs(title = "Clusters") +
  Seurat::NoAxes() + Seurat::NoLegend() +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5))

# Cluster type
plot_list[[4]] = Seurat::DimPlot(sobj, group.by = "cluster_type",
                                 reduction = name2D,
                                 cols = color_markers) +
  ggplot2::labs(title = "Cluster type") +
  Seurat::NoAxes() +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5),
                 legend.position = "none")

patchwork::wrap_plots(plot_list, nrow = 2)
```

We keep tumor cells based on cluster annotation.

```{r is_of_interest}
sobj$is_of_interest = (sobj$cluster_type == "tumor cells")
```


## Subset dataset

```{r subset_sobj}
sobj = subset(sobj, is_of_interest)
sobj
```
We clean the dataset :


```{r diet_sobj}
sobj = Seurat::DietSeurat(sobj)
sobj
```

How many cells by sample ?

```{r table_orig_ident}
table(sobj$project_name)
```

We represent this information as a barplot :

```{r barplot_count, fig.width = 4, fig.height = 5}
aquarius::plot_barplot(df = table(sobj$project_name) %>%
                         as.data.frame.table() %>%
                         dplyr::mutate(x = "Cells"),
                       x = "x", y = "Freq", fill = "Var1",
                       position = position_fill()) +
  ggplot2::scale_fill_manual(values = sample_info$color,
                             breaks = sample_info$project_name,
                             name = "Sample of origin")
```


# Processing

## Normalization

We remove genes that are expressed in less than 5 cells :

```{r filter_genes}
sobj = aquarius::filter_features(sobj, min_cells = 5)
sobj
```

We normalize gene expression :

```{r normalization}
sobj = Seurat::NormalizeData(sobj,
                             normalization.method = "LogNormalize")
sobj = Seurat::FindVariableFeatures(sobj, nfeatures = 2000)
sobj = Seurat::ScaleData(sobj)

sobj
```

## Non-corrected projection

We perform a PCA :

```{r pca}
sobj = Seurat::RunPCA(sobj,
                      assay = "RNA",
                      reduction.name = "RNA_pca",
                      npcs = 100,
                      seed.use = 1337L)
sobj
```

We choose the number of dimensions such that they summarize 60 % of the variability :

```{r ndims}
stdev = sobj@reductions[["RNA_pca"]]@stdev
stdev_prop = cumsum(stdev)/sum(stdev)
ndims = which(stdev_prop > 0.60)[1]
ndims
```

We can visualize this on the elbow plot :

```{r elbowplot, fig.width = 12, fig.height = 4}
elbow_p = Seurat::ElbowPlot(sobj, ndims = 100, reduction = "RNA_pca") +
  ggplot2::geom_point(x = ndims, y = stdev[ndims], col = "red")
x_text = ggplot_build(elbow_p)$layout$panel_params[[1]]$x$get_labels() %>% as.numeric()
elbow_p = elbow_p +
  ggplot2::scale_x_continuous(breaks = sort(c(x_text, ndims)), limits = c(0, 100))
x_color = ifelse(ggplot_build(elbow_p)$layout$panel_params[[1]]$x$get_labels() %>%
                   as.numeric() %>% round(., 2) == round(ndims, 2), "red", "black")
elbow_p = elbow_p +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = element_text(color = x_color))

elbow_p
```

We generate a tSNE and a UMAP with `r ndims` principal components :

```{r tsne_umap}
sobj = Seurat::RunTSNE(sobj,
                       reduction = "RNA_pca",
                       dims = 1:ndims,
                       seed.use = 1337L,
                       num_threads = n_threads, # Rtsne::Rtsne option
                       reduction.name = paste0("RNA_pca_", ndims, "_tsne"))

sobj = Seurat::RunUMAP(sobj,
                       reduction = "RNA_pca",
                       dims = 1:ndims,
                       seed.use = 1337L,
                       reduction.name = paste0("RNA_pca_", ndims, "_umap"))
```

We can visualize the two representations. There is a batch effect :

```{r see_umap_tsne, fig.width = 8, fig.height = 4, class.source = "fold-hide"}
tsne = Seurat::DimPlot(sobj, group.by = "project_name",
                       reduction = paste0("RNA_pca_", ndims, "_tsne")) +
  ggplot2::scale_color_manual(values = sample_info$color,
                              breaks = sample_info$project_name) +
  Seurat::NoAxes() + ggplot2::ggtitle("PCA - tSNE") +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5),
                 legend.position = "none")

umap = Seurat::DimPlot(sobj, group.by = "project_name",
                       reduction = paste0("RNA_pca_", ndims, "_umap")) +
  ggplot2::scale_color_manual(values = sample_info$color,
                              breaks = sample_info$project_name) +
  Seurat::NoAxes() + ggplot2::ggtitle("PCA - UMAP") +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5))

tsne | umap
```


## Batch-effect correction

We remove sample-specific effect with FastMNN :

```{r fastmnn}
sobj_mnn = Seurat::SplitObject(sobj,
                               split.by = "project_name")
sobj_mnn = aquarius::integration_fastmnn(object.list = sobj_mnn,
                                         assay = "RNA",
                                         verbose = TRUE,
                                         d = 100)

sobj@assays[["mnn.reconstructed"]] = sobj_mnn@assays[["mnn.reconstructed"]]
sobj@reductions[["mnn"]] = sobj_mnn@reductions[["mnn"]]
reduction_name = "mnn"

rm(sobj_mnn)

sobj
```

## Clustering

We make a clustering on the batch-effect corrected space :

```{r clustering}
sobj = Seurat::FindNeighbors(sobj, reduction = reduction_name)
sobj = Seurat::FindClusters(sobj, resolution = 1)
```

## UMAP and tSNE

To visualize subpopulations, we run UMAP and tSNE :

```{r mnn_tsne_umap}
sobj = Seurat::RunTSNE(sobj,
                       reduction = reduction_name,
                       dims = 1:ndims,
                       seed.use = 1337L,
                       num_threads = n_threads, # Rtsne::Rtsne option
                       reduction.name = paste0(reduction_name, "_", ndims, "_tsne"))

sobj = Seurat::RunUMAP(sobj,
                       reduction = reduction_name,
                       dims = 1:ndims,
                       seed.use = 1337L,
                       reduction.name = paste0(reduction_name, "_", ndims, "_umap"))
```

We can visualize the two representations :

```{r see_mnn_umap_tsne, fig.width = 8, fig.height = 4}
tsne = Seurat::DimPlot(sobj, group.by = "project_name",
                       reduction = paste0(reduction_name, "_", ndims, "_tsne")) +
  ggplot2::scale_color_manual(values = sample_info$color,
                              breaks = sample_info$project_name) +
  Seurat::NoAxes() + ggplot2::ggtitle("MNN - tSNE") +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5),
                 legend.position = "none")

umap = Seurat::DimPlot(sobj, group.by = "project_name",
                       reduction = paste0(reduction_name, "_", ndims, "_umap")) +
  ggplot2::scale_color_manual(values = sample_info$color,
                              breaks = sample_info$project_name) +
  Seurat::NoAxes() + ggplot2::ggtitle("MNN - UMAP") +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5))

tsne | umap
```


## Trajectory projection

Just for the visualization, we generate a diffusion map :

```{r dm, fig.width = 8, fig.height = 8}
sobj = aquarius::run_diffusion_map(sobj = sobj,
                                   input = reduction_name,
                                   seed = 1337L,
                                   verbose = TRUE,
                                   n_eigs = 50,
                                   suppress_dpt = TRUE,
                                   return_dm = FALSE)

dm_name = paste0(reduction_name, "_dm")

Seurat::DimPlot(sobj, reduction = dm_name,
                dims = c(2,4),
                group.by = "orig.ident") +
  ggplot2::scale_color_manual(values = sample_info$color,
                              breaks = sample_info$sample_identifiant) +
  ggplot2::labs(title = dm_name) +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
  Seurat::NoAxes()
```

We generate a UMAP from this diffusion map :

```{r dm_umap}
umap_dm_dims = 16

if (paste0(dm_name, "_umap") %in% names(sobj@reductions)) {
  sobj[[paste0(dm_name, "_umap")]] = NULL
}
sobj = Seurat::RunUMAP(sobj, reduction = dm_name,
                       dims = 1:umap_dm_dims,
                       reduction.name = paste0(dm_name, "_umap"),
                       verbose = TRUE, seed.use = 1337L)
```

We represent sample of origin on this projection :

```{r dm_umap_visu, fig.width = 8, fig.height = 8}
Seurat::DimPlot(sobj, reduction = paste0(dm_name, "_umap"),
                group.by = "project_name") +
  ggplot2::scale_color_manual(values = sample_info$color,
                              breaks = sample_info$sample_identifiant) +
  ggplot2::labs(title = paste0(dm_name, "_umap")) +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
  Seurat::NoAxes()
```


# Visualization

We use this projection :

```{r name2D}
name2D = paste0(reduction_name, "_", ndims, "_umap")
```

This is the clustering :

```{r see_clusters, fig.width = 8, fig.height = 8}
Seurat::DimPlot(sobj, reduction = name2D,
                group.by = "seurat_clusters", label = TRUE) +
  ggplot2::labs(title = name2D) +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
  Seurat::NoAxes()
```

This is the trajectory UMAP, colored by sample of origin :

```{r project_name, fig.width = 8, fig.height = 8}
Seurat::DimPlot(sobj, reduction = name2D,
                group.by = "project_name") +
  ggplot2::scale_color_manual(values = sample_info$color,
                              breaks = sample_info$sample_identifiant) +
  ggplot2::labs(title = name2D) +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
  Seurat::NoAxes()
```

Where are cells from each dataset ?

```{r tumor_split_by_origin, fig.width = 12, fig.height = 7}
# Build plots
plot_list = aquarius::plot_split_dimred(
  sobj = sobj,
  reduction = name2D,
  split_by = "project_name",
  split_color = setNames(nm = sample_info$project_name,
                         sample_info$color))

# Patchwork
patchwork::wrap_plots(plot_list, nrow = 3) &
  ggplot2::theme(legend.position = "none")
```

We represent XIST expression levels :

```{r xist_split_by_origin, fig.width = 12, fig.height = 7}
# Build plots
plot_list = aquarius::plot_split_dimred(
  sobj = sobj,
  reduction = paste0(reduction_name, "_", ndims, "_umap"),
  split_by = "project_name",
  split_color = setNames(nm = sample_info$project_name,
                         sample_info$color),
  color_by = "XIST")

# Patchwork
patchwork::wrap_plots(plot_list, nrow = 3) &
  ggplot2::theme(legend.position = "none")
```


# Save

We save the Seurat object :

```{r save_sobj}
saveRDS(sobj, file = paste0(out_dir, "/", save_name, "_sobj.rds"))
```


# R Session

```{r sessioninfo, echo = FALSE, fold_output = TRUE}
sessionInfo()
```

