---
title: "Figures from Brunet *et al.*"
author: "Audrey"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    code_folding: show
    code_download: true
    toc: true
    toc_float: true
    number_sections: false
---

<style>
body {
text-align: justify}
</style>

<!-- Automatically computes and prints in the output the running time for any code chunk -->
```{r, echo=FALSE}
# https://github.com/rstudio/rmarkdown/issues/1453
hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    res = hooks[[type]](x, options)
    
    if (isFALSE(options[[paste0("fold_", type)]])) return(res)
    
    paste0(
      "<details><summary>", "show", "</summary>\n\n",
      res,
      "\n\n</details>"
    )
  }
}
knitr::knit_hooks$set(
  output = hook_foldable("output"),
  plot = hook_foldable("plot"),
  time_it = local({
    now = NULL
    function(before, options) {
      if (options$time_it) {
        if (before) {
          now <<- Sys.time()
        } else {
          res = difftime(Sys.time(), now, units = "secs")
          paste("(Time to run :", round(res, digits = 2), "s)")
        }
      }
    }
  })
)
```

<!-- Set default parameters for all chunks -->
```{r, setup, include = FALSE}
set.seed(1337L)
knitr::opts_chunk$set(echo = TRUE, # display code
                      # display chunk output
                      message = FALSE,
                      warning = FALSE,
                      fold_output = FALSE, # usefull for sessionInfo()
                      fold_plot = FALSE,
                      
                      # figure settings
                      fig.align = 'center',
                      fig.width = 20,
                      fig.height = 15,
                      
                      # something about seed, chunk and Rmarkdown compilation
                      # https://stackoverflow.com/questions/39417003/long-vectors-not-supported-yet-error-in-rmd-but-not-in-r-script
                      # cache = TRUE,
                      cache.lazy = FALSE, 
                      
                      # add runtime after chunk
                      time_it = FALSE,
                      
                      # save figures in PDF in a separate folder
                      # https://stackoverflow.com/questions/27992239/knitr-include-figures-in-report-and-output-figures-to-separate-files
                      dev = c('png', 'pdf'), # tiff or pdf alone renders bad in html
                      # dpi = 300,
                      fig.path = "figures_detail/",
                      pdf.options(encoding = "ISOLatin9.enc"))
```


In this file, we make the scRNAseq figures for the paper.

```{r library}
library(dplyr)
library(patchwork)
library(ggplot2)
library(ComplexHeatmap)

.libPaths()
```

# Settings

Analyses are stored here :

```{r locations}
analysis_dir = "./../"

list.files(analysis_dir)
```

Output files will be stored there :

```{r out_dir}
out_dir = "./data/"

if (!dir.exists(out_dir)) {
  dir.create(out_dir)
}
```

## Annotations

These are all the samples analysed :

```{r figZ_1, class.source = 'fold-hide', fig.width = 10, fig.height = 7}
sample_info = readRDS(paste0(analysis_dir, "/1_metadata/patients_sample_info.rds"))
plot_list = aquarius::fig_plot_gb(sample_info, title = "NF1 patients datasets")

# patchwork
patchwork::wrap_plots(plot_list) +
  patchwork::plot_layout(nrow = length(plot_list), heights = c(0.1,4)) &
  ggplot2::theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15))
```

We load custom colors for cell types :

```{r color_markers, fig.width = 7, fig.height = 7}
color_markers = readRDS(paste0(analysis_dir, "/1_metadata/patients_color_markers.rds"))

graphics::pie(rep(1, length(color_markers)),
              col = unlist(color_markers),
              labels = names(color_markers))
```

We load gene name to assess cell type annotation on a dotplot :

```{r dotplot_markers}
dotplot_markers = readRDS(paste0(analysis_dir, "/1_metadata/patients_dotplot_markers.rds"))

lengths(dotplot_markers)
```

We define custom colors for tumor type :

```{r tumor_type_colors}
tumor_type_colors = c("pNF" = "#1c55b8",
                      "dyNF" = "#95c595",
                      "MPNST" = "#ad4fc5")
```

We load the list of cell markers to save as CSV for supplementary materials :

```{r cell_markers}
cell_markers = readRDS(paste0(analysis_dir, "/1_metadata/patients_cell_markers.rds"))
max_nrows = max(lengths(cell_markers))
cell_markers_df = lapply(c(1:length(cell_markers)), FUN = function(index_i) {
  vec = cell_markers[[index_i]] %>%
    sort()
  vec = c(vec, rep(NA, max_nrows - length(vec)))
  
  df = vec %>%
    as.data.frame()
  colnames(df) = names(cell_markers)[index_i]
  
  return(df)
}) %>%
  do.call(cbind, .)

head(cell_markers_df)
```


## Metadata

In this section, we represent all the available information about patients. Information are coming from :

* [Sun et al.](https://doi.org/10.1016/j.stem.2021.04.029) : materials and methods
* [Wu et al.](https://doi.org/10.1126/sciadv.abo5442) : supplementary table S3

```{r metadata_figure, fig.width = 12, fig.height = 6}
sample_info_ht = sample_info %>%
  dplyr::mutate(tumor_type = as.character(tumor_type)) %>%
  dplyr::mutate(tumor_type = factor(tumor_type, levels = c("nNF", "pNF", "dyNF", "MPNST", "metastasis"))) %>%
  dplyr::mutate(CDKN2A = case_when(project_name %in% c("GSM5404036", "GSM5404044") ~ "CN loss",
                                   project_name %in% c("GSM5404038", "GSM5404041", "2020_12") ~ "normal",
                                   project_name %in% c("2019_L", "2020_11") ~ "del hetero",
                                   project_name %in% c("2020_21") ~ "del homo")) %>%
  dplyr::mutate(TP53 = case_when(project_name %in% c("GSM5404036", "GSM5404044",
                                                     "GSM5404038", "GSM5404041",
                                                     "2019_L", "2020_11", "2020_21") ~ "normal",
                                 project_name %in% c("2020_12") ~ "del hetero")) %>%
  dplyr::mutate(PRC2 = case_when(project_name %in% c("GSM5404036", "GSM5404044",
                                                     "GSM5404038", "GSM5404041",
                                                     "2019_L", "2020_11") ~ "normal",
                                 project_name %in% c("2020_12") ~ "del hetero SUZ12",
                                 project_name %in% c("2020_21") ~ "del hetero EED + mut 43% SUZ12")) %>%
  dplyr::mutate(chemotherapy = case_when(
    project_name == "GSM5404035" ~ "Ifos/Dox 3 mo. earlier",
    project_name == "GSM5404036" ~ "Ifos/Dox 4 mo. earlier + Ifos/Etop 2 mo. earlier",
    project_name == "GSM5404040" ~ "Ifos/Dox",
    project_name == "GSM5404042" ~ "Ifos/Dox 3 mo. earlier",
    project_name == "GSM5404043" ~ "Ifos/Dox 3 mo. earlier",
    project_name == "GSM5404044" ~ "Ifos/Dox 4 mo. earlier")) %>%
  dplyr::mutate(chemotherapy = ifelse(is.na(chemotherapy) & origin %in% c("here", "Wu"),
                                      yes = "No", no = chemotherapy)) %>%
  dplyr::mutate(radiation = ifelse(project_name == "GSM5404035",
                                   yes = "2 mo. earlier", no = NA)) %>%
  dplyr::mutate(radiation = ifelse(is.na(radiation) & origin %in% c("here", "Wu"),
                                   yes = "No", no = radiation)) %>%
  dplyr::mutate(age = case_when(
    project_name %in% c("GSE183308_1", "GSE183308_2", "GSE183308_3", "GSE183308_4") ~ 18,
    project_name == "GSM5050721" ~ 19,
    project_name == "2019_L" ~ 53,
    project_name == "2020_11" ~ 22,
    project_name == "2020_12" ~ 50,
    project_name == "2020_21" ~ 25)) %>%
  dplyr::arrange(tumor_type, origin, gender)

# Heatmap colors
list_colors = list()

list_colors[["project_name"]] = setNames(sample_info_ht$color,
                                         nm = sample_info_ht$project_name)
list_colors[["tumor_type"]] = tumor_type_colors
list_colors[["gender"]] = setNames(c("#FF9900", "#32CD32"),
                                   nm = c("F", "M"))
list_colors[["origin"]] = setNames(aquarius::gg_color_hue(length(unique(sample_info_ht$origin))),
                                   nm = unique(sample_info_ht$origin))
list_colors[["chemotherapy"]] = setNames(RColorBrewer::brewer.pal(
  n = length(na.omit(unique(sample_info_ht$chemotherapy))), name = "Dark2"),
  nm = na.omit(unique(sample_info_ht$chemotherapy)))
list_colors[["chemotherapy"]][names(list_colors[["chemotherapy"]]) == "No"] = "lightgray"

list_colors[["radiation"]] = setNames(RColorBrewer::brewer.pal(
  n = length(na.omit(unique(sample_info_ht$radiation))), name = "Dark2")[c(1:2)],
  nm = na.omit(unique(sample_info_ht$radiation)))
list_colors[["radiation"]][names(list_colors[["radiation"]]) == "No"] = "lightgray"

gene_status = sample_info_ht[, c("CDKN2A", "TP53", "PRC2")] %>%
  unlist() %>% na.omit() %>% unique()
list_colors[["gene_status"]] = setNames(aquarius::c30_palette[c(1:length(gene_status))],
                                        nm = gene_status)
list_colors[["gene_status"]][names(list_colors[["gene_status"]]) == "normal"] = "lightgray"

# Heatmap annotation
ha = ComplexHeatmap::HeatmapAnnotation(
  project_name = sample_info_ht$project_name,
  tumor_type = sample_info_ht$tumor_type,
  origin = sample_info_ht$origin,
  gender = sample_info_ht$gender,
  age = anno_points(sample_info_ht$age, height = unit(2, "cm")),
  CDKN2A = sample_info_ht$CDKN2A,
  TP53 = sample_info_ht$TP53,
  PRC2 = sample_info_ht$PRC2,
  chemotherapy = sample_info_ht$chemotherapy,
  radiation = sample_info_ht$radiation,
  col = list(project_name = list_colors[["project_name"]],
             tumor_type = list_colors[["tumor_type"]],
             gender = list_colors[["gender"]],
             origin = list_colors[["origin"]],
             CDKN2A = list_colors[["gene_status"]],
             TP53 = list_colors[["gene_status"]],
             PRC2 = list_colors[["gene_status"]],
             chemotherapy = list_colors[["chemotherapy"]],
             radiation = list_colors[["radiation"]]),
  show_legend = c(FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE),
  annotation_label = c("color", "tumor type", "origin", "gender (from XIST)",
                       "age", "CDKN2A", "TP53", "PRC2", "chemotherapy", "radiation"),
  na_col = "white")

# Matrix
empty_mat = matrix(nrow = 1, ncol = nrow(sample_info_ht), data = 0)
colnames(empty_mat) = sample_info_ht$project_name

# Heatmap
ht = ComplexHeatmap::Heatmap(empty_mat,
                             top_annotation = ha,
                             cluster_rows = FALSE,
                             cluster_columns = FALSE,
                             height = unit(0, "cm"),
                             column_names_rot = 40,
                             column_names_side = "top")

# Draw
ComplexHeatmap::draw(ht,
                     show_heatmap_legend = FALSE,
                     legend_grouping = "original",
                     annotation_legend_side = "bottom")
```

## Dataset

We load the human dataset.

```{r load_human_dataset}
sobj = readRDS(paste0(analysis_dir, "/3_combined/all_patients_sobj.rds"))
sobj
```

This is the projection of interest to visualize cells on :

```{r name2D}
name2D = "harmony_80_tsne"
```

We change the levels of cell types:

```{r custom_levels_cell_type}
custom_levels = c("tumor cells", "fibroblasts", "endothelial cells", "mural cells",
                  "macrophages", "cDC", "T cells", "NK cells", "mast cells", "B cells")

# In sobj
sobj$cell_type = factor(as.character(sobj$cell_type),
                        levels = custom_levels)
table(sobj$cell_type, useNA = "ifany")

# In color_markers
color_markers = color_markers[custom_levels]

# In dotplot_markers
dotplot_markers = dotplot_markers[custom_levels]
```

We group cell type by cluster, to correct mis-annotation :

```{r cluster_type}
cluster_type = table(sobj$cell_type, sobj$seurat_clusters) %>%
  prop.table(., margin = 2) %>%
  apply(., 2, which.max)
cluster_type = setNames(nm = names(cluster_type),
                        levels(sobj$cell_type)[cluster_type])

sobj$cluster_type = cluster_type[sobj$seurat_clusters]
sobj$cluster_type = factor(sobj$cluster_type,
                           levels = levels(sobj$cell_type))

table(sobj$cluster_type)
```


# Figures

## Annotation

In this section, we visualize cell type annotation.

### Single-cell level

We look at the annotation :

```{r see_cell_type_single, fig.width = 7, fig.height = 6}
fig_cell_type = Seurat::DimPlot(sobj, group.by = "cell_type",
                                reduction = name2D, cols = color_markers) +
  Seurat::NoAxes() +
  ggplot2::labs(title = "Cell Type",
                subtitle = "single cell level annotation") +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5),
                 plot.subtitle = element_text(hjust = 0.5),
                 legend.position = "right")
fig_cell_type
```

We look at the annotation, split by sample :

```{r see_cell_type_split_single, fig.width = 15, fig.height = 10}
plot_list = aquarius::plot_split_dimred(sobj,
                                        reduction = name2D,
                                        split_by = "project_name",
                                        split_color = sample_info$color,
                                        group_by = "cell_type",
                                        group_color = color_markers)

patchwork::wrap_plots(plot_list, nrow = 3) &
  Seurat::NoLegend()
```

### Cluster level

We visualize the clustering :

```{r see_cluster, fig.width = 7, fig.height = 6}
Seurat::DimPlot(sobj, group.by = "seurat_clusters",
                reduction = name2D, label = TRUE) +
  Seurat::NoAxes() +
  ggplot2::labs(title = "Clusters",
                subtitle = paste0(ncol(sobj), " cells")) +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5),
                 plot.subtitle = element_text(hjust = 0.5),
                 legend.position = "none")
```

This is the annotation smoothed by cluster :

```{r see_cell_type_cluster, fig.width = 7, fig.height = 6}
fig_cell_type = Seurat::DimPlot(sobj, group.by = "cluster_type",
                                reduction = name2D, cols = color_markers) +
  Seurat::NoAxes() +
  ggplot2::labs(title = "Cell Type",
                subtitle = "cluster level annotation") +
  ggplot2::theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5),
                 plot.subtitle = element_text(hjust = 0.5),
                 legend.position = "right")
fig_cell_type
```

and split by sample :

```{r see_cell_type_split_cluster, fig.width = 15, fig.height = 10}
plot_list = aquarius::plot_split_dimred(sobj,
                                        reduction = name2D,
                                        split_by = "project_name",
                                        split_color = sample_info$color,
                                        group_by = "cluster_type",
                                        group_color = color_markers)

patchwork::wrap_plots(plot_list, nrow = 3) &
  Seurat::NoLegend()
```

## By project name

We look at `project_name` annotation :

```{r see_project_name, fig.width = 7, fig.height = 6}
# Random order
set.seed(1234)
rnd_order = sample(colnames(sobj), replace = FALSE, size = ncol(sobj))

# Extract coordinates
cells_coord = sobj@reductions[[name2D]]@cell.embeddings %>%
  as.data.frame() %>%
  `colnames<-`(c("Dim1", "Dim2"))
cells_coord$project_name = sobj$project_name
cells_coord$tumor_type = sobj$tumor_type
cells_coord = cells_coord[(rnd_order), ]

# Plot
fig_project_name = ggplot2::ggplot(cells_coord, aes(x = Dim1, y = Dim2, col = project_name)) +
  ggplot2::geom_point(size = 0.1) +
  ggplot2::scale_color_manual(values = sample_info$color,
                              breaks = sample_info$project_name) +
  ggplot2::labs(title = "Project Name") +
  ggplot2::theme_void() +
  ggplot2::theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
                 aspect.ratio = 1) +
  guides(color  = guide_legend(title = "Project Name",
                               override.aes = list(size = 3)))
fig_project_name
```

We look at `tumor_type` annotation :

```{r see_tumor_type, fig.width = 7, fig.height = 6}
fig_tumor_type = ggplot2::ggplot(cells_coord, aes(x = Dim1, y = Dim2, col = tumor_type)) +
  ggplot2::geom_point(size = 0.1) +
  ggplot2::scale_color_manual(values = tumor_type_colors,
                              breaks = names(tumor_type_colors)) +
  ggplot2::labs(title = "Tumor Type") +
  ggplot2::theme_void() +
  ggplot2::theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
                 aspect.ratio = 1) +
  guides(color  = guide_legend(title = "Tumor Type",
                               override.aes = list(size = 3)))
fig_tumor_type
```

We represent cell type split by tumor type :

```{r cell_type_split, fig.width = 12, fig.height = 5}
plot_list = aquarius::plot_split_dimred(sobj,
                                        reduction = name2D,
                                        split_by = "tumor_type",
                                        group_by = "cluster_type",
                                        group_color = unlist(color_markers))

fig_split = patchwork::wrap_plots(plot_list, nrow = 1) &
  Seurat::NoLegend()
  
fig_split
```

We look at sample composition :

```{r barplot, fig.width = 12, fig.height = 6}
sobj$project_name = sobj$project_name %>%
  as.character() %>%
  factor(., levels = sample_info %>%
           dplyr::arrange(tumor_type, as.character(project_name)) %>%
           dplyr::pull(project_name) %>% as.character())

aquarius::plot_barplot(df = table(sobj$project_name,
                                  sobj$cluster_type) %>%
                         as.data.frame.table() %>%
                         `colnames<-`(c("Sample", "Cell Type", "Number")),
                       x = "Sample", y = "Number", fill = "Cell Type",
                       position = position_stack()) +
  ggplot2::scale_fill_manual(values = unlist(color_markers),
                             breaks = names(color_markers),
                             name = "Cell Type") +
  ggplot2::theme(axis.title = element_blank(),
                 axis.text.x = element_text(color = sample_info$color))
```

## Quality Control

This is the log number of UMI detected in each dataset :

```{r vln_log_nCount_RNA, fig.width = 12, fig.height = 7}
Seurat::VlnPlot(sobj, features = "log_nCount_RNA", pt.size = 0.01,
                group.by = "project_name", cols = sample_info$color) +
  ggplot2::theme(axis.title.x = element_blank()) +
  ggplot2::guides(fill = guide_legend(ncol = 1))
```

This is the number of features detected in each dataset :

```{r vln_nFeature_RNA, fig.width = 12, fig.height = 7}
Seurat::VlnPlot(sobj, features = "nFeature_RNA", pt.size = 0.01,
                group.by = "project_name", cols = sample_info$color) +
  ggplot2::theme(axis.title.x = element_blank()) +
  ggplot2::guides(fill = guide_legend(ncol = 1))
```

This is the percentage of UMI related to genes from mitochondrial genome, detected in each dataset :

```{r vln_percent.mt, fig.width = 12, fig.height = 7}
Seurat::VlnPlot(sobj, features = "percent.mt", pt.size = 0.01,
                group.by = "project_name", cols = sample_info$color) +
  ggplot2::theme(axis.title.x = element_blank()) +
  ggplot2::guides(fill = guide_legend(ncol = 1))
```

This is the percentage of UMI related to ribosomal subunits, detected in each dataset :

```{r vln_percent.rb, fig.width = 12, fig.height = 7}
Seurat::VlnPlot(sobj, features = "percent.rb", pt.size = 0.01,
                group.by = "project_name", cols = sample_info$color) +
  ggplot2::theme(axis.title.x = element_blank()) +
  ggplot2::guides(fill = guide_legend(ncol = 1))
```

This is the XIST expression level, used to distinguish women (XIST+) from men (XIST-) :

```{r vln_xist, fig.width = 12, fig.height = 7}
Seurat::VlnPlot(sobj, features = "XIST", pt.size = 0.01,
                group.by = "project_name", cols = sample_info$color) +
  ggplot2::theme(axis.title.x = element_blank()) +
  ggplot2::guides(fill = guide_legend(ncol = 1))
```

This dotplot assesses the cell type annotation :

```{r fig6_Y, class.source = 'fold-hide', fig.width = 8, fig.height = 5}
markers = dotplot_markers[levels(sobj$cluster_type)] # new order
markers = unique(unlist(markers)) # as vector
markers = markers[!(markers %in% c("PTPRC"))]
markers = c("PTPRC", markers)
markers = markers[markers %in% rownames(sobj)]

fig_dotplot = aquarius::fig_plot_dotplot(sobj,
                                         groupby = "cluster_type",
                                         markers = markers,
                                         gene_expression_colors = aquarius::color_gene)
fig_dotplot
```

We represent expression levels for those markers on the projection :

```{r see_markers, fig.width = 12, fig.height = 13}
plot_list = lapply(markers, FUN = function(one_gene) {
  p = Seurat::FeaturePlot(sobj, features = one_gene,
                          pt.size = 0.001,
                          reduction = name2D, order = TRUE) +
    ggplot2::scale_color_gradientn(colors = aquarius::color_gene) +
    Seurat::NoAxes() +
    ggplot2::theme(aspect.ratio = 1,
                   plot.title = element_text(hjust = 0.5),
                   legend.position = "right")
  
  return(p)
})
names(plot_list) = markers

patchwork::wrap_plots(plot_list, ncol = 4)
```

## Dotplot

We define genes of interest, to represent on the dotplot :

```{r genes_of_interest}
genes_of_interest = c("TIGIT", "VSIR", "LGALS9", "HAVCR2", "SIRPA", "CD47",
                      "PDCD1", "ICOS", "LAG3", "CD274", "PDCD1LG2",
                      "FAS", "FASLG", "GZMB", "GZMH", "GZMA", "GZMK",
                      "PRF1", "GNLY", "IDO1", "IL4I1", "ARG2", "NOS2",
                      "HLA-A", "HLA-B", "HLA-C", 
                      # RAS-related
                      "MAP2K1", "MAP2K2", "MAPK1", "MAPK3", "CCND1",  # MAPK pathway
                      "KRAS", "NRAS", "HRAS", "ARAF", "BRAF", "RAF1") # RAS and RAF

```

We make the figure :

```{r dotplot, fig.width = 13, fig.height = 12}
plot_list = lapply(levels(sobj$cluster_type), FUN = function(one_cell_type) {
  subsobj = subset(sobj, cluster_type == one_cell_type)
  
  p = Seurat::DotPlot(subsobj, assay = "RNA",
                      features = genes_of_interest,
                      group.by = "tumor_type",
                      scale = TRUE,
                      scale.by = "radius") +
    ggplot2::labs(title = one_cell_type) +
    ggplot2::scale_y_discrete(position = "right") +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 0),
                   plot.title = element_text(hjust = 0.5, size = 10),
                   axis.title = element_blank(),
                   axis.text = element_text(size = 10)) +
    ggplot2::scale_color_gradientn(colors = aquarius::color_cnv) +
    ggplot2::coord_flip()
  
  return(p)
})

p1 = plot_list[[1]]
plot_list = lapply(plot_list, FUN = function(one_plot) {
  axis_text = ggplot_build(one_plot)$layout$panel_scales_y[[1]]$range$range
  axis_color = tumor_type_colors[axis_text]
  
  one_plot +
    ggplot2::theme(axis.text.y = element_blank(),
                   axis.line.y = element_blank(),
                   axis.ticks.y = element_blank(),
                   legend.position = "none",
                   # axis.text.x = ggplot2::element_text(angle = 0, hjust = 0.5, size = 20,
                   #                                     color = axis_color),
                   text = element_text(size = 15)) #+
  # ggplot2::scale_y_discrete(labels = rep("\U25CF", 3),
  #                           position = "right")
})
plot_list[[1]] = p1 +
  ggplot2::theme(legend.position = "left")

fig_main_dotplot = patchwork::wrap_plots(plot_list, nrow = 1) +
  patchwork::plot_layout(guides = "collect")
fig_main_dotplot
```


## Violin plot

In this section, we make violin plot within specific cell types. So, we set cells identifier as cell type smoothed by cluster :

```{r change_idents}
Seurat::Idents(sobj) = sobj$cluster_type
table(Seurat::Idents(sobj))
```

### T cells

Violin plot within T cells :

```{r vln_t_cells_sample, fig.width = 12, fig.height = 7}
genes_of_interest = c("SIRPA", "HAVCR2", "PDCD1")

plot_list = lapply(genes_of_interest, FUN = function(one_gene) {
  p = Seurat::VlnPlot(sobj, features = one_gene, pt.size = 0.01,
                      idents = "T cells",
                      group.by = "project_name", cols = sample_info$color) +
    ggplot2::theme(axis.title.x = element_blank()) +
    ggplot2::guides(fill = guide_legend(ncol = 1))
  
  return(p)
})
names(plot_list) = genes_of_interest

plot_list
```

### Macrophages

Violin plot within macrophages :

```{r vln_macrophages_sample, fig.width = 12, fig.height = 7}
genes_of_interest = c("SIRPA", "HAVCR2", "PDCD1", "CCND1")

plot_list = lapply(genes_of_interest, FUN = function(one_gene) {
  p = Seurat::VlnPlot(sobj, features = one_gene, pt.size = 0.01,
                      idents = "macrophages",
                      group.by = "project_name", cols = sample_info$color) +
    ggplot2::theme(axis.title.x = element_blank()) +
    ggplot2::guides(fill = guide_legend(ncol = 1))
  
  return(p)
})
names(plot_list) = genes_of_interest

plot_list
```


Same figure split by tumor type :

```{r vln_macrophages_tumor_type, fig.width = 6, fig.height = 7}
plot_list = lapply(genes_of_interest, FUN = function(one_gene) {
  p = Seurat::VlnPlot(sobj, features = one_gene, pt.size = 0.01,
                      idents = "macrophages",
                      group.by = "tumor_type", cols = tumor_type_colors) +
    ggplot2::theme(axis.title.x = element_blank()) +
    ggplot2::guides(fill = guide_legend(ncol = 1)) +
    Seurat::NoLegend()
  
  return(p)
})
names(plot_list) = genes_of_interest

plot_list
```

### Tumor cells

Violin plot within tumor cells :

```{r vln_tumor_cells_sample, fig.width = 12, fig.height = 7}
genes_of_interest = c("HLA-A", "HLA-B", "HLA-C")

plot_list = lapply(genes_of_interest, FUN = function(one_gene) {
  p = Seurat::VlnPlot(sobj, features = one_gene, pt.size = 0.01,
                      idents = "tumor cells",
                      group.by = "project_name", cols = sample_info$color) +
    ggplot2::theme(axis.title.x = element_blank()) +
    ggplot2::guides(fill = guide_legend(ncol = 1))
  
  return(p)
})
names(plot_list) = genes_of_interest

plot_list
```


Same figure split by tumor type :

```{r vln_tumor_cells_tumor_type, fig.width = 6, fig.height = 7}
plot_list = lapply(genes_of_interest, FUN = function(one_gene) {
  p = Seurat::VlnPlot(sobj, features = one_gene, pt.size = 0.01,
                      idents = "tumor cells",
                      group.by = "tumor_type", cols = tumor_type_colors) +
    ggplot2::theme(axis.title.x = element_blank()) +
    ggplot2::guides(fill = guide_legend(ncol = 1)) +
    Seurat::NoLegend()
  
  return(p)
})
names(plot_list) = genes_of_interest

plot_list
```

## Assembly suggestion

In this section, we suggest figures assemblies.

### Main figure

Figure 1 - **Single-cell RNA-sequencing from 18 NF1 patients**. (A) Representation showing 78,518 cells on a tSNE, colored according to tumor type. (B) Representation split by tumor type and colored by cell type. Number of cells associated with each tumor type is written on top of each tSNE. Gray background enables to compare cells location between each tSNE. (C) Dotplot showing gene expression split by cell type and by tumor type. Dot size represents proportion of cells expressing corresponding marker. Dot color represents average gene expression levels for cells concerned by the dot.


```{r fig_main, fig.width = 13, fig.height = 13}
plot_list = list(fig_tumor_type,
                 fig_split +
                   patchwork::plot_layout(guides = "collect") &
                   ggplot2::theme(legend.position = "right") &
                   guides(color  = guide_legend(title = "Cell type",
                                                override.aes = list(size = 3))),
                 fig_main_dotplot)

patchwork::wrap_plots(plot_list, design = "AB\nCC", heights = c(1, 2.5), widths = c(1, 3)) +
  patchwork::plot_annotation(title = "Main Figure",
                             theme = theme(plot.title = element_text(hjust = 0.5, size = 20)),
                             tag_levels = list(c("A", "B", "", "", "C"))) &
  theme(plot.tag = element_text(face = "plain", size = 20))
```

### Supplementary figure

Figure S1 - **Single-cell RNA-sequencing from 18 NF1 patients**. (A) Representation showing 78,518 cells on a tSNE, colored according to cell type (a) or project name (b). (B) Dotplot assessing cell populations annotation. Dot size represents proportion of cells expressing corresponding marker. Dot color represents average gene expression levels for cells concerned by the dot.

```{r fig_supp, fig.width = 13, fig.height = 10}
plot_list = list(fig_cell_type +
                   guides(color  = guide_legend(title = "Cell type",
                                                override.aes = list(size = 3))),
                 fig_project_name,
                 fig_dotplot)

patchwork::wrap_plots(plot_list, design = "AB\nCC") +
  patchwork::plot_annotation(title = "Supplementary Figure",
                             theme = theme(plot.title = element_text(hjust = 0.5, size = 20)),
                             tag_levels = list(c("A   a", "b", "", "", "B"))) &
  theme(plot.tag = element_text(face = "plain", size = 20))
```

Alternative layout and tag :

```{r fig_supp2, fig.width = 13, fig.height = 10}
plot_list = list(fig_project_name,
                 fig_cell_type,
                 fig_dotplot)

patchwork::wrap_plots(plot_list, design = "AC\nBC") +
  patchwork::plot_annotation(title = "Supplementary Figure",
                             theme = theme(plot.title = element_text(hjust = 0.5, size = 20)),
                             tag_levels = 'A') &
  theme(plot.tag = element_text(face = "plain", size = 20))
```

# References

- **GSE165826** : [https://doi.org/10.1016/j.stem.2021.04.029](https://doi.org/10.1016/j.stem.2021.04.029)
- **GSE179033** : [https://doi.org/10.1126/sciadv.abo5442](https://doi.org/10.1126/sciadv.abo5442)
- **RStudio** : RStudio Team (2020). RStudio: Integrated Development for R. RStudio, PBC, Boston, MA URL http://www.rstudio.com/
- **rmarkdown** : Yihui Xie and J.J. Allaire and Garrett Grolemund (2018). R Markdown: The Definitive Guide. Chapman and Hall/CRC. ISBN 9781138359338. URL https://bookdown.org/yihui/rmarkdown.
- **Singularity** : [https://doi.org/10.1371/journal.pone.0177459](https://doi.org/10.1371/journal.pone.0177459)
- **Seurat V3** : [https://doi.org/10.1016/j.cell.2019.05.031](https://doi.org/10.1016/j.cell.2019.05.031)
- **scDblFinder** : [https://doi.org/10.12688/f1000research.73600.2](https://doi.org/10.12688/f1000research.73600.2)
- **scds** : [https://doi.org/10.1093/bioinformatics/btz698](https://doi.org/10.1093/bioinformatics/btz698)
- **harmony** : [https://doi.org/10.1038/s41592-019-0619-0](https://doi.org/10.1038/s41592-019-0619-0)
- **ggplot2** : Wickham H (2016). ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York. ISBN 978-3-319-24277-4, https://ggplot2.tidyverse.org.
- **patchwork** : Pedersen T (2022). patchwork: The Composer of Plots. https://patchwork.data-imaginist.com, https://github.com/thomasp85/patchwork.
- **ComplexHeatmap** : [https://doi.org/10.1093/bioinformatics/btw313](https://doi.org/10.1093/bioinformatics/btw313)


# Save

We save the gene-markers used for cell type annotation, for Supplementary Materials:

```{r cell_markers_df}
write.table(cell_markers_df, file = paste0(out_dir, "/cell_markers.csv"),
            col.names = TRUE, na = "", row.names = FALSE)
```


# R Session

```{r sessioninfo, echo = FALSE, fold_output = TRUE}
sessionInfo()
```

