---
title: "NF1 patients"
subtitle: "Make metadata"
author: "Audrey"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    code_folding: show
    code_download: true
    toc: true
    toc_float: true
    number_sections: false
---

<style>
body {
text-align: justify}
</style>

<!-- Automatically computes and prints in the output the running time for any code chunk -->
```{r, echo=FALSE}
# https://github.com/rstudio/rmarkdown/issues/1453
hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    res = hooks[[type]](x, options)
    
    if (isFALSE(options[[paste0("fold_", type)]])) return(res)
    
    paste0(
      "<details><summary>", "show", "</summary>\n\n",
      res,
      "\n\n</details>"
    )
  }
}
knitr::knit_hooks$set(
  output = hook_foldable("output"),
  plot = hook_foldable("plot"),
  time_it = local({
    now = NULL
    function(before, options) {
      if (options$time_it) {
        if (before) {
          now <<- Sys.time()
        } else {
          res = difftime(Sys.time(), now, units = "secs")
          paste("(Time to run :", round(res, digits = 2), "s)")
        }
      }
    }
  })
)
```

<!-- Set default parameters for all chunks -->
```{r, setup, include = FALSE}
set.seed(1337L)
knitr::opts_chunk$set(echo = TRUE, # display code
                      # display chunk output
                      message = FALSE,
                      warning = FALSE,
                      fold_output = FALSE, # usefull for sessionInfo()
                      fold_plot = FALSE,
                      
                      # figure settings
                      fig.align = 'center',
                      fig.width = 20,
                      fig.height = 15,
                      
                      # something about seed, chunk and Rmarkdown compilation
                      # https://stackoverflow.com/questions/39417003/long-vectors-not-supported-yet-error-in-rmd-but-not-in-r-script
                      # cache = TRUE,
                      cache.lazy = FALSE, 
                      
                      # add runtime after chunk
                      time_it = FALSE)
```


This file is used to generate :

* **sample_info** : dataframe containing sample names, sample path, custom colors...
* **color_markers** : named list to give custom colors to cell types
* **cell_markers** : named list containing markers for each cell type

```{r library}
library(dplyr)
library(patchwork)
library(ggplot2)

.libPaths()
```


# Preparation

In this section, we set the global settings of the analysis. We will store data there :

```{r out_dir}
save_name = "patients"
out_dir = "."
```

# Sample info

We create a dataframe containing everything except colors.

```{r sample_info}
sample_info = data.frame(
  project_name = c("2019_L", "2020_11", "2020_12", "2020_21",
                   "GSM5404032", "GSM5404033", "GSM5404034",                    # Wu
                   "GSM5404035", "GSM5404036", "GSM5404037", "GSM5404038",      # Wu
                   "GSM5404039", "GSM5404040", "GSM5404041", "GSM5404042",      # Wu
                   "GSM5404043", "GSM5404044",                                  # Wu
                   "GSM5050721"),                                              # Sun
  genotype = "patient",
  tumor_type = c(rep("dyNF", 3), "MPNST",
                 rep("pNF", 9), rep("MPNST", 4),
                 "dyNF"),
  gender = c("F", rep("M", 3),
             "F", "M", "F",
             "F", "M", "M", "F",
             "F", "M", "M", "M",
             "F", "M",
             "M"),
  color = NA,
  origin = c(rep("here", 4),
             rep("Wu", 13),
             rep("Sun", 1)),
  stringsAsFactors = FALSE) %>%
  dplyr::mutate(sample_type = paste0(genotype, "_", tumor_type)) %>%
  dplyr::mutate(tumor_type = factor(tumor_type, levels = c("pNF", "dyNF", "MPNST"))) %>%
  dplyr::arrange(genotype, tumor_type, gender) %>%
  dplyr::mutate(project_name = factor(project_name, levels = project_name)) %>%
  dplyr::mutate(unique_id = data.table::rowid(sample_type)) %>%
  dplyr::mutate(sample_identifier = paste0(sample_type, "_", unique_id)) %>%
  dplyr::mutate(sample_identifier = factor(sample_identifier, levels = sample_identifier)) %>%
  dplyr::select(project_name,  genotype, tumor_type, gender,
                origin, sample_identifier, sample_type, color)

sample_info
```

We add custom colors for each sample :

```{r custom_palette}
list_palette = list(
  # patients
  patient_pNF = colorRampPalette(c("dodgerblue", "midnightblue")),
  patient_dyNF = colorRampPalette(c("darkseagreen1", "darkseagreen4")),
  patient_MPNST = colorRampPalette(c("mediumorchid1", "mediumorchid4")))

for (sample_type in unique(sample_info$sample_type)) {
  matching_dataset = (sample_info$sample_type == sample_type)
  sample_info[matching_dataset, "color"] = list_palette[[sample_type]](sum(matching_dataset))
}
```


We visualize the colors :

```{r custom_palette_see, fig.width = 8, fig.height = 8}
graphics::pie(rep(1, nrow(sample_info)),
              col = sample_info$color,
              labels = sample_info$project_name)
```

We save it :

```{r save_sample_info}
saveRDS(sample_info, file = paste0(out_dir, "/", save_name, "_sample_info.rds"))
```


# Cell types

We define markers for each population of interest :

```{r cell_markers, class.source = "fold-hide"}
cell_markers = list(
  "tumor cells" = c("ANK3", "PLP1", "NAV2", "COL18A1", "POSTN", "SOX4", "SERPINE2", "GPM6B",
                    "CALD1", "PDGFA", "TIMP3", "DST", "FBLN2", "LGI4", "SERPINH1",
                    "NR2F2", "ITGB8", "COL5A2", "PTPRZ1", "LAMA4", "GJC3", "TENM3", "COL1A2",
                    "ZNF536", "COL16A1", "DLGAP1", "STARD13", "COL27A1", "SOX10",
                    "TANC2", "ERBB3", "PDZD2", "COL1A1", "COL3A1", "SPARC", "PDLIM4", "EMP1",
                    "DOCK7", "KLF9", "HS2ST1", "NEDD4", "GAS7", "UTRN", "SPECC1", "FRMD4A",
                    "TPM1", "DUSP6", "GAS1", "CXCL12", "PDGFRB", "COL6A1", "HTRA1", "LOXL1",
                    "MAST4", "SLIT2", "COL4A2", "PCOLCE", "CPE", "DKK2", "CAV1", "PRSS23",
                    "TEAD1", "GJA1", "WWTR1", "AEBP1", "NREP", "NID1", "ITGBL1",
                    "ZFHX4", "FBN1", "PCDH19", "DDR2", "SDC2", "MXRA7", "CNN3",
                    "LTBP1", "FOXG1", "LTBP4", "LHFP", "RBFOX2", "TPM2", "RYK", "GPX8",
                    "PXDN", "RCN1", "FKBP9", "KCNA1", "CHL1", "AATK", "KCNA2", "CADM4",
                    "ART3", "CDH19", "CD59", "MAL", "CADM1", "ATP1A2", "PTN", "GATM",
                    "CNP", "CRYAB", "VWA1", "EMP2", "PRNP", "POU3F1", "MBP", "SCN7A",
                    "CKB", "SCD", "MATN2", "HSPG2", "FABP5", "ABCA8", "DLK1"),
  
  "macrophages" = c("C1QC", "MAFB", "FCGR3A", "FCGR3B", "TREM2", "CX3CR1", "TMEM176A",
                    "CYBB", "FCGR1A", "ADGRE1", "TF", "ITGB5",
                    "TLR2", "CEBPA", "C3AR1", "PLA2G7", "ZNF705A", "SIRPA", "CD68", "ITGAM",
                    "LAIR1", "PYHIN1", "CYP4F3", "NCF2", "C1QA", "PID1", "SLC15A3" , "DAB2",
                    "CLEC12A", "HCK", "LRP1", "LILRB3", "CCR1",
                    "LILRB4", "RASSF4", "P2RX4", "GNGT2", "SYK", "IRF5", "LYZ"),
  
  "cDC" = c("FCER1A", "CLEC10A", "CD1C", "FCGR2B", "IL1R2", "LGALS2", "CD1E", "CD1D", "AMICA1",
            "PKIB", "RGS18", "ITGAX", "DAPP1", "GAPT", "IGSF6", "CIITA", "HLA-DOA", "PLD4",
            "STX11", "HCK", "ZNF385A", "DOK2"),
  
  "T cells" = c("CD3E", "CD8A", "CD4", "CD3G", "NKG7",
                "CXCR6", "MS4A4A", "PTPRCAP", "GIMAP4", "SH2D2A", "CD28", "CD2",
                "LCK", "LAT", "BCL11B", "THY1", "GIMAP6", "ITK", "GIMAP1", "SKAP1", "IL2RB",
                "RHOH", "SEPT1"),
  
  "NK cells" = c("KLRB1", "NCR1", "STYK1", "XCL1", "KLRC2", "CA2",
                 "SAMD3", "SATB1", "KLRK1", "KLRD1", "TXK",
                 "CD244", "CD244", "STAT4", "FASLG", "GIMAP5",
                 "GNLY", "NKG7", "PRF1", "GZMB", "XCL2", "HOPX", "GZMA", "CD247"),
  
  "mast cells" = c("MRGPRX2", "TPSB2", "TPSAB1", "FCER1A", "CPA3",
                   "SLC6A4", "CMA1", "KIT", "RAB27B",
                   "GATA2", "IL1RL1", "HDC", "SERPINB1",
                   "LAT2", "VWA5A", "CCL7", "TMEM64",
                   "BASP1", "CCL2", "GCH1", "FNIP1", "LILRB4"),
  
  "B cells" = c("CD79A", "LY6D", "MS4A1", "BANK1", "FCMR", "CCR7", "CD79B", "HLA-DOB", "BCL11A",
                "GIMAP6", "HLA-DOA", "PTPRCAP", "CD37", "HLA-DMB", "LTB", "EBF1", "SCD", "NAPSA",
                "CYTIP", "REL", "POU2F2", "MZB1", "FCRL5", "DERL3", "IGJ", "IGLL5", "PIM2",
                "FAM46C", "TXNDC11"),
  
  "endothelial cells" = c("EGFL7", "CDH5", "EMCN", "CD34", "ESAM", "PLVAP", "APLNR", "RAMP2", "TSHZ2",
                          "SOX18", "ECSCR", "CAV1", "ADGRF5", "PTPRB", "RASIP1", "TIE1", "KDR", "CYYR1",
                          "MMRN2", "FAM110D", "FLT1", "ARHGAP29", "LRG1", "SPARCL1", "VWF", "KANK3",
                          "AQP1", "COL4A2", "COL4A1", "NFIB", "ENG", "MAST4", "S100A16", "CRIP2", "PECAM1",
                          "ABCG2", "CD200", "APBB2", "HSPG2", "GNG11", "IGFBP7", "PRKCDBP", "CAVIN3", "MCAM",
                          "TM4SF1", "PDLIM1", "INSR", "SMAD1", "CD93", "PLEKHG1", "SERPINH1", "SPARC", "IGFBP4",
                          "PPIC", "PLXND1", "COL18A1", "SOX4", "TIMP3", "SNRK", "DST", "EMP1"),
  
  "fibroblasts" = c("MFAP5", "LUM", "FBLN1", "SERPING1", "CILP", "ASPN", "CDH11", "MFAP4",
                    "CCDC80", "SVEP1", "LOX", "IL33", "LOXL1", "CPXM1", "MFAP2", "FBN2",
                    "ISLR", "C1S", "CPXM2", "CLEC11A", "FNDC1", "CD248", "PDPN", "SLIT3",
                    "FIBIN", "COLEC12", "TWIST1", "PDGFRL", "GXYLT2", "PDGFRA", "IGSF10", "CTHRC1",
                    "BICC1", "SNHG18", "FKBP11", "CTSK", "DCN", "ADAMTS2", "FZD2", "MMP23A",
                    "MDK", "SMOC2", "SCARF2", "EBF1", "KDELR3", "FBN1", "IGF1", "C1R",
                    "GPX3", "MRC2", "P3H3", "MGP", "PCOLCE", "CDC42EP5", "ZFHX4", "COL12A1",
                    "COL14A1", "PRRX1", "RBP1", "RHOJ", "LHFP", "GPX8", "COL8A1", "MXRA8",
                    "COL6A2", "EFEMP2", "COL6A3", "SCN7A", "APOD", "FZD1", "LPAR1", "DPYSL3",
                    "PLPP3", "FKBP7", "AEBP1", "SLIT2", "PALLD", "BMP1", "IGFBP6", "FSTL1",
                    "GAS6", "FKBP10", "MAGED2", "CRTAP", "RBMS3", "COL5A1", "PLOD2", "OAF",
                    "MMP2", "GRB10", "COPZ2", "SERPINF1", "PLAT", "TIMP1", "FKBP9", "COL5A3",
                    "S100A16", "RAB34", "SERF1A", "SERF1B", "THBS2", "NPDC1", "MATN2", "FAM114A1", "FSCN1",
                    "MEG3", "MGST1", "ECM1", "COL16A1", "DDAH2"),
  
  "mural cells" = c("DES", "MYH11", "ANO1", "NDUFA4L2", "RGS5", "NOTCH3", "GUCY1A3", "GUCY1A1", "PDE3A",
                    "MYLK", "TPM2", "NRARP", "APOLD1", "MAP3K20", "C11orf96", "TINAGL1", "MYL9",
                    "ARHGEF17", "USP2", "RASD1", "MYO1B", "ASPN", "MUSTN1", "S1PR3", "PTP4A3", "SLC25A25",
                    "PDGFRB", "PPP1R12B", "RASL11A", "RBPMS", "MFGE8", "ADAMTS9", "MGST3", "RASGRP2", "RRAD",
                    "GNG11", "CRISPLD2", "TPM1", "EPAS1", "MPRIP", "HSPB1", "ZFHX3", "BICD2", "SYNE2",
                    "CWC25", "CRTC3", "NOLC1", "LPP", "DLC1", "MAPRE2", "CYGB", "MAFK", "NUDT4",
                    "TNS1", "IRS2", "THY1", "CPE", "RBM4B", "FBXO33", "ARHGEF12"))

lengths(cell_markers)
```

Here are custom colors for each cell type :

```{r color_markers, fig.width = 8, fig.height = 0.8, class.source = "fold-hide"}
color_markers = aquarius::color_markers
color_markers = color_markers[names(color_markers) %in% names(cell_markers)]
cell_markers = cell_markers[names(color_markers)] # re-order

data.frame(cell_type = names(color_markers),
           color = unlist(color_markers)) %>%
  ggplot2::ggplot(., aes(x = cell_type, y = 0, fill = cell_type)) +
  ggplot2::geom_point(pch = 21, size = 5) +
  ggplot2::scale_fill_manual(values = unlist(color_markers), breaks = names(color_markers)) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "none",
                 axis.line = element_blank(),
                 axis.title = element_blank(),
                 axis.ticks = element_blank(),
                 axis.text.y = element_blank(),
                 axis.text.x = element_text(angle = 10))
```

These are some markers to assess cell type annotation :

```{r dotplot_markers}
dotplot_markers = list(
  "tumor cells" = c("PLP1", "SOX10", "SOX9", "DLK1"),
  "macrophages" = c("LYZ", "ADGRE1"),
  "cDC" = c("CD1C", "Itgae"),
  "T cells" = c("CD3E", "CD8A", "CD4"),
  "NK cells" = c("KLRB1C", "NCR1"),
  "mast cells" = c("TPSB2", "KIT"),
  "B cells" = c("CD79A", "CD19"),
  "endothelial cells" = c("EGFL7", "PECAM1"),
  "fibroblasts" = c("MFAP5", "LUM"),
  "mural cells" = c("NOTCH3", "DES"))

lengths(dotplot_markers)
```

We save them :

```{r save_annotation}
saveRDS(color_markers, file = paste0(out_dir, "/", save_name, "_color_markers.rds"))
saveRDS(cell_markers, file = paste0(out_dir, "/", save_name, "_cell_markers.rds"))
saveRDS(dotplot_markers, file = paste0(out_dir, "/", save_name, "_dotplot_markers.rds"))
```


# R Session

```{r sessioninfo, echo = FALSE, fold_output = TRUE}
sessionInfo()
```
